Mental Health Factors and Covid

### 2025-04-20 products/manuscript/Manuscript.qmd
Part 5 working on manuscript

# Summary/Abstract
_Write a summary of your project._

@@

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

@@

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

There's 151406 observations and 47 variables.

@@

The following code details the data cleaning process. Labels are removed from the data to make it easier to work with to create models later in the process. Variables needed are selected, then a series of binary varibales are created. Additional details are available in the processing folder.
```{r}
here() #set working directory
data_original <- readRDS(here("data","raw-data", "NHIS_COVID.rds"))
data1 <- labelled::remove_val_labels(data_original) #removes labels from haven package which cause problems

data1 <- data_original %>%
  select(STRATA, PSU, PERNUM, SAMPWEIGHT, AGE, PA18AERSTR, HRSLEEP, DEPFREQ, PHQSLEEP, CMPSUPPORT, CVDTESTRSLT)

data2 <- data1 %>%
  mutate(
    PA18AERSTR = ifelse(PA18AERSTR %in% c(0, 7, 8, 9), NA, PA18AERSTR), #put NA, 0, 6, 7, 8, 9 to NA
    EXERCISE = if_else(PA18AERSTR %in% c(2,3,4), 1, 0, missing = NA_real_) #create EXERCISE variable, 1 if guidelines by Am Heart Assoc met for aerobics, strength, or both, 0 if not met
    ) %>%
    mutate(
    AGE = ifelse(AGE %in% c(997, 999), NA, AGE),  #invalid ages to NA
    SLEEP = case_when( #Based on age, amount of sleep required from https://www.nhlbi.nih.gov/health/sleep/how-much-sleep#:~:text=Experts%20recommend%20that%20adults%20sleep,or%20more%20hours%20a%20night.
      AGE < 1  ~ if_else(HRSLEEP >= 12 & HRSLEEP <= 16, 1, 0, missing = NA_real_),
      AGE >= 1 & AGE < 3       ~ if_else(HRSLEEP >= 11 & HRSLEEP <= 14, 1, 0, missing = NA_real_),
      AGE >= 3 & AGE < 6       ~ if_else(HRSLEEP >= 10 & HRSLEEP <= 13, 1, 0, missing = NA_real_),
      AGE >= 6 & AGE < 13      ~ if_else(HRSLEEP >= 9 & HRSLEEP <= 12, 1, 0, missing = NA_real_),
      AGE >= 13 & AGE <= 18    ~ if_else(HRSLEEP >= 8 & HRSLEEP <= 10, 1, 0, missing = NA_real_),
      TRUE                     ~ NA_real_  #default NA for unspecified ages
    )) %>%
  mutate(
    DEPRESSED = case_when(
      DEPFREQ %in% c(1, 2) ~ 1, #daily and weekly is considered depressed
      DEPFREQ %in% c(3, 4, 5)    ~ 0, #monthly and a few times a year and never are considered not depressed
      DEPFREQ %in% c(0, 7, 8, 9) ~ NA_real_,#NIU & Unknowns are NA
      TRUE ~ NA_real_ #default NA for unexpected values
    )) %>%
  mutate(
    TROUBLE_SLEEPING = case_when(
      PHQSLEEP %in% c(1, 2, 3) ~ 1, #several days, more than half days, nearly every day are trouble sleeping
      PHQSLEEP == 0            ~ 0,  #not at all is no trouble sleeping
      PHQSLEEP %in% c(6, 7, 8, 9) ~ NA_real_,  #NIU & Unknowns are NA
      TRUE ~ NA_real_  #default NA for unexpected values
    )) %>%
  mutate(
    SOCIAL = case_when(
      CMPSUPPORT == 1 ~ 0,   #more social and emotional support
      CMPSUPPORT == 2 ~ 1,  #less social and emotional support
      CMPSUPPORT == 3 ~ 0,   #about the same
      CMPSUPPORT %in% c(0, 7, 8, 9) ~ NA_real_,  # NIU & Unknowns → NA
      TRUE ~ NA_real_  #default NA for unexpected values
    )) %>%
  mutate(
    COVID = case_when(
      CVDTESTRSLT == 2 ~ 1,  #yes, had COVID
      CVDTESTRSLT == 1 ~ 0,  #no, did not have COVID
      CVDTESTRSLT %in% c(0, 3, 7, 8, 9) ~ NA_real_,  #NIU, did not receive results, unknowns, and NA are NA
      TRUE ~ NA_real_  #default NA for unexpected values
    ))
```

@@

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@@

```{r}
data <- data2
#summarize EXERCISE
summary_exercise <- data %>%
  summarise(
    Variable = "EXERCISE",
    Prop_1 = mean(EXERCISE == 1, na.rm = TRUE),
    Prop_0 = mean(EXERCISE == 0, na.rm = TRUE),
    Prop_NA = mean(is.na(EXERCISE))
  )

#summarize SLEEP
summary_sleep <- data %>%
  summarise(
    Variable = "SLEEP",
    Prop_1 = mean(SLEEP == 1, na.rm = TRUE),
    Prop_0 = mean(SLEEP == 0, na.rm = TRUE),
    Prop_NA = mean(is.na(SLEEP))
  )

#summarize DEPRESSED
summary_depressed <- data %>%
  summarise(
    Variable = "DEPRESSED",
    Prop_1 = mean(DEPRESSED == 1, na.rm = TRUE),
    Prop_0 = mean(DEPRESSED == 0, na.rm = TRUE),
    Prop_NA = mean(is.na(DEPRESSED))
  )

#summarize TROUBLE SLEEPING
summary_trouble_sleeping <- data %>%
  summarise(
    Variable = "TROUBLE SLEEPING",
    Prop_1 = mean(TROUBLE_SLEEPING == 1, na.rm = TRUE),
    Prop_0 = mean(TROUBLE_SLEEPING == 0, na.rm = TRUE),
    Prop_NA = mean(is.na(TROUBLE_SLEEPING))
  )

#summarize SOCIAL
summary_social <- data %>%
  summarise(
    Variable = "SOCIAL",
    Prop_1 = mean(SOCIAL == 1, na.rm = TRUE),
    Prop_0 = mean(SOCIAL == 0, na.rm = TRUE),
    Prop_NA = mean(is.na(SOCIAL))
  )

#summarize COVID
summary_covid <- data %>%
  summarise(
    Variable = "COVID",
    Prop_1 = mean(COVID == 1, na.rm = TRUE),
    Prop_0 = mean(COVID == 0, na.rm = TRUE),
    Prop_NA = mean(is.na(COVID))
  )

#combine all summaries and convert to gt table
summary_table <- bind_rows(
  summary_exercise, 
  summary_sleep, 
  summary_depressed, 
  summary_trouble_sleeping, 
  summary_social, 
  summary_covid
) %>%
  gt() %>%
  tab_header(
    title = "Proportions of 1, 0, and NA for Various Variables"
  )

#print the combined table
summary_table

#list variables
varstable1 = c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", 
             "SOCIAL", "COVID")

#list categories
cat_vartab1 = c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", 
             "SOCIAL", "COVID")

#use function to make table
Tab1 = CreateTableOne(vars = varstable1, factorVars = cat_vartab1, data = data)

# Convert the TableOne object to a data frame
tab1_df <- as.data.frame(print(Tab1, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE))

# Clean up row names
tab1_df <- tibble::rownames_to_column(tab1_df, "Variable")

# More advanced formatting
tab1_gt <- gt(tab1_df) %>%
  # Add a title
  tab_header(
    title = md("**COVID test status and Mental Health Factors**"),
    subtitle = "Survey Results Summary"
  ) %>%
  # Format the columns
  fmt_percent(
    columns = contains("%"),
    decimals = 1
  ) %>%
  # Group the variables
  tab_row_group(
    label = "COVID Status",
    rows = contains("COVID")
  ) %>%
  tab_row_group(
    label = "Mental Health Indicators",
    rows = contains(c("DEPRESSED", "TROUBLE_SLEEPING"))
  ) %>%
  tab_row_group(
    label = "Lifestyle Factors",
    rows = contains(c("EXERCISE", "SLEEP", "SOCIAL"))
  ) %>%
  # Style the table
  tab_style(
    style = cell_fill(color = "#e6f3ff"),
    locations = cells_row_groups()
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_row_groups()
  ) %>%
  # Add footnotes if needed
  tab_footnote(
    footnote = "Percentages may not sum to 100% due to rounding",
    locations = cells_column_labels(columns = contains("%"))
  ) %>%
  # Adjust column widths
  cols_width(
    Variable ~ px(250),
    everything() ~ px(120)
  )

tab1_gt
```

@@

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

>>>>

# Summary/Abstract

Isolation and quarantine strategies employed for containment of SARS-Cov-2 infection led to unexpected mental health struggles among the population of the United States. Infected individuals and healthy individuals were affected, but little is known about the potential relationships between mental health and physical health. In this study, we utilized Nataional Health Information Surveys from 2019-2021 to examine the relationships between receieving a positive COVID-19 test and self reporting mental health factors such as meeting recommended guidelines for exercise, meeting the recommended number of hours of sleep per night, reporting feeling depressed recently, reporting having trouble sleeping recently, and reporting receiving less social support.

We find

@@

@@

Data was subsetted to years 2019-2022, to cover the brunt of pandemic years. Despite low prevelance of COVID-19 cases in 2019 worldwide, the appearance of COVID-19 in the news (and consequential potential impact on mental health factors) led us to include this data. The final dataset contains 151406 observations.

Variables of interest contained significant amounts of missing information, so multiple imputation was used, creating 5 sets of imputed data that was averaged across the remainder of analysis to produce variables and statistics needed. Complex survey design was used to allow for generalizability of results, using vairbales PSU, STRATA, and SAMPWIEGHT.

Binary variables were created to allow for interpretability.

~~~

@@

The following code details the data cleaning process. Labels are removed from the data to make it easier to work with to create models later in the process. Variables needed are selected, then a series of binary varibales are created. Additional details are available in the processing folder.

@@

@@

@@
